\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-04-24 forwards-and-backwards.tex
alison edits 4/22

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. The vast majority of these
results concern spin systems \comment{TODO REFs obtained from Flandoli/Huang etc: Alison}.
However,~\cite{flandoli/huang:2021} 
obtain the Fisher-KPP equation
as a scaling limit of interacting diffusions in
$\IR^d$. In that setting a major challenge (also apparent in our models), not
faced in spin systems, is 
the control of concentration of particles. We also note that the scaling 
considered in~\cite{flandoli/huang:2021} is different from that considered
below. In their language we consider only `moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas they also consider the situation in which
that number remains finite;
furthermore we shall consider a scaling in which 
the reproduction mechanism is near-critical and look at large timescales,
while \citet{flandoli/huang:2021} only scale the number of
particles in the system and the size of the neighbourhood over which indivduals interact.

@@

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', 
\citep{birzu/hallatschek/korolev:2017, birzu/hallatschek/korolev:2021}.
In that setting, the genealogies will be quite different
from the Fisher-KPP case. 
%\comment{Could refer to more recent Birzu et al paper here.}
For example, \citet{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/ex2a/pme_1234.steps}
        \includegraphics[width=0.7\textwidth]{figures/ex2b/pme_1234.steps}
    \end{center}
    \caption{
        Simulated 1D populations under a porous medium equation with logistic growth~(\ref{PME}) in $d=1$,
        % comparing the simulated profile to the analytical solution;
        $\theta/N$ small on the left; large on the right.
        Simulation is with $\gamma(m) = m$ and $\mu(m) = (1+1/\theta) m - 1/\theta$
        (implemented by choosing a small $dt$ taking reproduction probability per time step
        to be $1 - \exp(- dt m)$
        and death probability to be $\mu(m) = 1 - \exp(-dt ((1+1/\theta) m - 1/\theta))$);
        values of $\theta$ are 1 and 100, respectively, and both values of $N$ are the same.
        \label{fig:pme_waves}
    }
\end{figure}

@@

Simulating these processes and exploring parameter space,
one sooner or later comes upon a strange observation:
with certain parameter combinations, the population spontaneously forms a regular grid
of more or less dicrete patches, separated by areas with nearly no individuals,
as shown in Figure~\ref{fig:clumping}.
The phenomenon is discussed in Section~16.10 of \citet{haller2022evolutionary},
and has been described in similar models, e.g.,
by \citet{britton1990spatial,sasaki1997clumped,hernandezgarcia2004clustering,young2001reproductive,berestycki2009nonlocal}.
Heuristically, this can occur if density-dependent effects of individuals
extend sufficiently farther than the typical dispersal distance:
the patches are stable because new offspring landing between two clumps
effectively compete with \emph{both} neighboring clumps,
while individuals within a clump effectively compete with only one.

The implication here is that the equilibrium solution to~\eqref{general deterministic limit}
may not be unique,
and a stable solution may have oscillations on a scale determined by the interaction distance.
We can see this easily in the simple case that
$\DG = \sigma^2 \Delta$,
the interaction kernels $\rho_r$, $\rho_\gamma$, and $\rho_F$
are all equal to $p_{\epsilon^2}$ (the heat kernel at time $\epsilon^2$),
and all parameters are spatially homogeneous,
so that $r(x, \eta) = r(\smooth{r}\eta(x))$, and similarly for $\gamma$ and $F$.
Let $\varphi_0$ be a value so that $F(\varphi_0) = 0$ and $F'(\varphi_0) < 0$,
so that we would expect $\varphi(x) \equiv \varphi_0$ to be a stable equilibrium solution.
Let us linearize~\eqref{general deterministic limit} about $\varphi_0$:
let $\varphi_t(x) = \varphi_0 + \psi_t(x)$,
and (informally) $r(x) \approx r(\varphi_0) + r'(\varphi_0) \smooth{r} \psi(x)$.
Writing $r_0 = r(\varphi_0)$ and $r'_0 = r'(\varphi_0)$,
and similarly for $\gamma$ and $F$,
\begin{align*}
    \partial_t \psi
    &\approx
    \sigma^2 \varphi_0 r_0 \gamma_0' \Delta \smooth{\gamma} \psi
    + \sigma^2 r_0 \gamma_0 \Delta \psi
    + \varphi_0 F'_0 \smooth{F} \psi .
\end{align*}
Letting $\hat \psi(u) = \int e^{2\pi i u x} \psi(x) dx$,
\begin{align*}
    \partial_t \hat \psi(u)
    &\approx
    \left(
        - \sigma^2 \varphi_0 r_0 \gamma_0' u^2 e^{-\epsilon^2 u^2 /2}
        - \sigma^2 r_0 \gamma_0 u^2
        + \varphi_0 F'_0 e^{-\epsilon^2 u^2 /2} 
    \right) \hat \psi(u) .
\end{align*}
Since $F'_0 < 0$, the last two terms are negative,
but on the other hand if we take $\gamma_0' < 0$
(i.e., per capita birth rate decreases with density near equilibrium),
then the stability of the constant solution $\varphi \equiv \varphi_0$
depends on the balance between the first and latter two terms.
Writing
$A_1 = - \sigma^2 \varphi_0 r_0 \gamma'_0$,
$A_2 = \sigma^2 r_0 \gamma_0$,
and $A_3 = - \varphi_0 F'_0$
(with signs chosen so each is positive),
and defining
$$ g(v) = A_1 v - A_2 v e^{\epsilon^2 v /2} - A_3 , $$
we have $\partial_t \hat \psi(u) = g(u^2) e^{\epsilon^2 u^2 /2} \psi(u)$.
The term $g(v)$ is maximized at $v_*$ solving
$e^{\epsilon^2 v_*/2} = A_1 / (A_2 (\epsilon^2 v_*/ 2 + 1))$,
which is positive if $A_1 > A_2$,
i.e., if $-\varphi_0 \gamma_0' > \gamma_0$.
If so, then the constant solution is unstable with respect to perturbations
that oscillate with frequency around $\sqrt{v_*}$.
\comment{This seems wrong, since it doesn't involve $\sigma$?
    But at least note the unstable frequencies scale with $\epsilon$.
    We might say more by linearizing the exponential term,
    which gives $v_* = (A_1 - A_2 + A_3 \epsilon^2/2)/(A_1 \epsilon^2)$,
    although the resulting expression isn't terribly helpful.
    TODO ALISON
}

@@

We now look at the second order term in~(\ref{generator prelimit}).
An individual at location $x$ gives birth 
to a surviving offspring at $y$ at rate
$$
\gamma(x,\eta) r(y,\eta) q_{\theta}(x, dy) ,
$$
and since this increments $\langle f, \eta \rangle$ by $f(y) / N$,
the contribution to the quadratic variation from birth events,
which occur at rate $\theta$ per individual 
(so, rate $N\theta |\eta|$ overall), is
$$
\int_{\IR^d}
    N \theta \gamma(x,\eta)
    \int \frac{1}{N^2} f^2(y) r(y,\eta)
    q_\theta(x,dy) 
\eta(dx) .
$$
Similarly, the increment in $\langle f, \eta\rangle$ resulting from 
the death of an individual at $x$ is $-f(x)/N$, and so combining with the 
above, the second order term in the generator takes the form
\begin{align*}
& G''(\langle f,\eta\rangle)
\frac{1}{2} N \theta
\left\{
    \int_{\IR^d}
        \gamma(x,\eta)
        \int \frac{1}{N^2}f^2(y)r(y,\eta)q_\theta(x,dy) 
    \eta(dx)
    +
    \int_{\IR^d}
        \mu_\theta(x,\eta)\frac{1}{N^2}f^2(x) 
    \eta(dx)
\right\} \\
&\qquad
= \frac{1}{2} G''(\langle f, \eta \rangle)
    \frac{\theta}{N}
    \int_{\IR^d}
        \gamma(x, \eta) \int f^2(y) r(y, \eta) q_\theta(x, dy) + f^2(x) \mu_\theta(x, \eta),
    \eta(dx) .
\end{align*}
Since $\int f^2(y) r(y, \eta) q_\theta(x, dy) \to f^2(x) r(x, \eta)$
and $r\gamma  + \mu_\theta = 2 r \gamma  - F / \theta \to 2 r \gamma$
as $\theta \to \infty$,
this converges to
\begin{align*}
\frac{\alpha}{2} G''(\langle f, \eta \rangle)
    \big\langle
	2 r(x, \eta) \gamma(x, \eta),
        \eta(dx)
    \big\rangle .
\end{align*}

@@

Putting these together,
the level $u$ of an individual at $x$ evolves according to:
\begin{equation} \label{eqn:dot_u}
    \dot u
    =
    - \theta\frac{u}{N} \left(N - u\right)
    \gamma(x, \eta) \int_{\IR^d} r(y, \eta) q_\theta(x, dy) 
    +
    \theta\mu_\theta(x,\eta) u .
\end{equation}
We shall write 
\begin{align} \label{b_defn}
    b_\theta(x, \eta)
    :=
    \theta\left(
    \gamma(x,\eta) \int_{\IR^d} r(y, \eta) q_\theta(x, dy)
    -
    \mu_\theta(x,\eta)
    \right) ,
\end{align}
which captures the local net difference between reproduction and death,
and
\begin{align} \label{c_defn}
    c_\theta(x, \eta)
    :=
    \frac{\theta}{N} \gamma(x, \eta) \int_{\IR^d} r(y, \eta) q_\theta(x, dy) ,
\end{align}
which captures the local rate of production of successful offspring.
Recall from equation~\eqref{eqn:mu_defn} that
$F(x,\eta) = \theta(r(x,\eta)\gamma(x,\eta) - \mu_\theta(x,\eta))$,
and so 
\begin{equation*} 
b_\theta(x, \eta)
=    \theta \gamma(x, \eta) \int_{\IR^d} \left( r(y, \eta) - r(x, \eta) \right) 
	q_\theta(x, dy)
    + F(x, \eta). 
\end{equation*}
Under Assumptions~\ref{def:model_setup}, as $\theta \to \infty$
$c_\theta(x, \eta)$ will tend to $\alpha \gamma(x, \eta) r(x, \eta)$
and $b_\theta(x, \eta)$ will tend to
\begin{equation}
\label{eqn:b_limit}
    \gamma(x, \eta) \DG r(x, \eta) + F(x, \eta) .
\end{equation}

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:lookdown_mgale}
For given positive values of $N$ and $\theta$, define the generator $A^N$
by
\begin{equation}
	\label{eqn:lookdown_generator}
\begin{split}
& A^{N}f(\lp ) \\
&\quad =
    f(\lp)
    \,\sum_{(x,u)\in \lp}\,
    2 \gamma(x, \eta)
    \Bigg\{ \frac{1}{2N} \int_u^N g(x,u_1) du_1
            \frac{
                \theta \int_{\IR^d}
                (g(y,u) - g(x,u))
                r(y, \eta) q_{\theta}(x,dy)
            }{ g(x,u) }
        \\
    &\qquad\qquad\qquad\qquad\qquad\qquad\qquad {} +
        \frac{\theta}{N} \int_u^N
        \int_{\IR^d}\left(
            \frac{ g(y,u_1) + g(x,u_1) }{ 2 } - 1
        \right)
        r(y, \eta) q_{\theta}(x,dy)
        du_1
    \Bigg\}\\
    &\qquad\qquad {} +
    f(\lp) \sum_{(x,u)\in\lp}\,
    \left(
        c_\theta(x,\eta) u^2 - b_{\theta}(x,\eta)u
    \right)
    \frac{\partial_u g(x,u)}{g(x,u)}
    ,
\end{split}
\end{equation}
where $f(\lp) = \prod_{(x, u) \in \lp} g(x, u)$ is as defined in \eqref{eqn:test_functions},
and $\eta(\cdot) = \lp(\cdot \times [0, N]) / N$
as before. 
Given $\lp_0 \in {\mathcal C}_N$, 
we say that a ${\mathcal D}_{[0,\infty)}({\mathcal C}_N)$-valued process 
$(\xi_t)_{t \ge 0}$
   is a solution to the $(A^N, \lp_0)$ martingale problem
    if 
$f(\lp_t) - f(\lp_0)-\int_0^t A^N f(\lp_s) ds$ is a martingale (with 
respect to the natural filtration) for all 
test functions $f$ 
    as defined above.
\end{definition}

@@

To make this precise,
define the spatial projection maps
$\kappa^N: \mathcal{M}(\IR^d \times [0, N]) \to \mathcal{M}(\IR^d)$
by $\kappa^N(\xi^N)(\cdot) = \xi^N(\cdot \times [0, N]) / N$,
and $\kappa: \mathcal{M}(\IR^d \times [0, \infty)) \to \mathcal{M}(\IR^d)$
by $\kappa(\xi)(\cdot) = \lim_{u_0 \to \infty} \xi(\cdot \times [0, u_0]) / u_0$.
We will also need an inverse notion:
for a measure $\lp^N$ on $\IR^d \times [0, N]$ and a sigma-field $\mathcal{F}$,
we say that \emph{$\lp^N$ is conditionally uniform given $\mathcal{F}$}
if for all compactly supported $f$,
\begin{align} \label{eqn:conditionally_uniform}
    \IE[ e^{-\langle f, \lp \rangle} \;|\; \mathcal{F} ]
    &=
    e^{-\langle H_f^N, \kappa^N(\lp) \rangle} ,
\end{align}
where
$$ H_f^N(x) = - N \log \frac{1}{N} \int_0^N e^{-f(x, u)} du . $$
In other words, the $[0, N]$ components of $\lp$
are independent, uniformly distributed on $[0, N]$, and independent of $\kappa(\lp)$.
Similarly, for a measure $\lp$ on $\IR^d \times [0, \infty)$
we say that \emph{$\lp$ is a conditionally Poisson random measure given $\mathcal{F}$}
if for all compactly supported $f$,
\begin{align} \label{eqn:conditionally_poisson}
    \IE[ e^{-\langle f, \lp \rangle} \;|\; \mathcal{F} ]
    &=
    e^{-\langle \int_0^\infty (1 - e^{-f(x, u)}) du, \kappa(\lp)(dx) \rangle} .
\end{align}
In other words, $\lp$ is conditionally Poisson with Cox measure $\kappa(\lp) \times \lambda$,
where $\lambda$ is Lesbegue measure.

>>>>

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. 
\comment{FROM ALISON: I rewrote the rest of this paragraph. Missing references
are in the latex file, next to the citation, but commented out}
Following the now classical contributions 
of~\citet{demasi/ferrari/lebowitz:1986, demasi/presutti:1991,
oeschlaeger:1985},
%@article{demasi/ferrari/lebowitz:1986,
%  title={Reaction-diffusion equations for interacting particle system},
%  author={De Masi, A. and Ferrari, P. A. and Lebowitz, J. L.},
%  journal={J. Stat. Phys.},
%  year={1986},
%  volume={44},
%  pages={589-644},
%  DOI = {10.1007/BF01011311},
%}
% A. De Masi and E. Presutti. Mathematical methods for hydrodynamic limits,
%volume 1501 of Lecture Notes in Mathematics. Springer-Verlag, Berlin, 1991.
%
% K. Oeschl\"ager. A law of large numbers for moderately interacting 
% diffusion processes. Z. Warsch. Verw. Gebiete. 69(2):279-322, 1985
much of this work has focused on lattice based models with one particle
per site, or on systems with a fixed number, $N$,
of interacting diffusions as $N\to\infty$.
For systems of proliferating particles,
as considered for example by 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019,
flandoli/huang:2021}, 
% K. Oeschl\"ager. On the derivation of reaction-diffusion equations as limits
% of systems of moderately interacting stochastic processes. Probab. Th. Rel.
% Fields 82, 565--586, (1989)
%
% F. Flandoli, M. Leimbach, and C. Olivera. Uniform convergence of 
%proliferating particles to the FKPP equation. J. Math. Anal. Appl. 473(1):
%27--52, (2019)
an additional challenge (also apparent in our models), is 
the control of concentration of particles. 
We follow 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019} in considering
`moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas 
\citet{flandoli/huang:2021} also consider the situation in which
that number remains finite. 
We refer to \citet{flandoli/huang:2021} for a more thorough 
literature review, 
but note that both our model and scaling differ from those considered
in the body of work discussed there: whereas in those settings, the only 
scalings are the number of particles in the system and the size of the
neighbourhood over which individuals interact with one another, in
keeping with the vast literature on continuous state branching models, we 
also scale time and so must ensure that births are 
adequately compensated by deaths to prevent the population from exploding.

@@

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', 
\citep{birzu/hallatschek/korolev:2017, birzu/hallatschek/korolev:2021}.
\comment{The missing reference is
 G. Birzu, O. Hallatshcek and K. Korolev. Genealogical structure
 changes as range expansions transition from pushed to pulled. 
 PNAS 118(34) (2021) 
 https://doi.org/10.1073/pnas.2026746118}
In that setting, the genealogies will be quite different
from the Fisher-KPP case. 
%\comment{Could refer to more recent Birzu et al paper here.}
For example, \citet{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/ex2a/pme_1234.steps}
        \includegraphics[width=0.7\textwidth]{figures/ex2b/pme_1234.steps}
    \end{center}
    \caption{
        Simulated 1D populations under a porous medium equation with logistic growth~(\ref{PME}) in $d=1$,
        % comparing the simulated profile to the analytical solution;
        $\theta/N$ small on the left; large on the right.
        Simulation is with $\gamma(m) = m$ and $\mu(m) = (1+1/\theta) m - 1/\theta$
        (implemented by choosing a small $dt$ taking reproduction probability per time step
        to be $1 - \exp(- dt m)$
        and death probability to be $\mu(m) = 1 - \exp(-dt ((1+1/\theta) m - 1/\theta))$);
        values of $\theta$ are 1 and 100, respectively, and both values of $N$ are the same.
	\comment{FROM ALISON Should we say what N is??}
        \label{fig:pme_waves}
    }
\end{figure}

@@

Simulating these processes and exploring parameter space,
one sooner or later comes upon a strange observation:
with certain parameter combinations, the population spontaneously forms a regular grid
of more or less dicrete patches, separated by areas with nearly no individuals,
as shown in Figure~\ref{fig:clumping}.
The phenomenon is discussed in Section~16.10 of \citet{haller2022evolutionary},
and has been described in similar models, e.g.,
by \citet{britton1990spatial,sasaki1997clumped,hernandezgarcia2004clustering,young2001reproductive,berestycki2009nonlocal}.
\comment{FROM ALISON: I rewrote the rest of this section - please check!}
These patches can be stable. For example, if
the density-dependent effects of individuals
extend farther, but not too much farther, than the typical dispersal distance,
new offspring landing between two clumps
can effectively find themselves in competition with 
\emph{both} neighbouring clumps,
while individuals within a clump compete with only one.

More mathematically, consider the case in which
$\DG = \sigma^2 \Delta$ and all 
parameters are spatially homogeneous,
so that $r(x, \eta) = r(\smooth{r}\eta(x))$, and similarly for $\gamma$ and $F$.
If $\varphi_0$ is such that $F(\varphi_0)=0$ and $F'(\varphi_0)<0$, then the
constant solution $\varphi\equiv\varphi_0$ is a nontrivial equilibrium
of~\eqref{general deterministic limit}. However, it may not be unique,
it may be unstable, and a stable solution may have 
oscillations on a scale determined by the interaction distance.

To understand the stability of the constant solution $\varphi\equiv\varphi_0$, 
we
linearise~\eqref{general deterministic limit} around $\varphi_0$:
let $\varphi_t(x) = \varphi_0 + \psi_t(x)$,
and (informally) $r(x) \approx r(\varphi_0) + r'(\varphi_0) \smooth{r} \psi(x)$.
Writing $r_0 = r(\varphi_0)$ and $r'_0 = r'(\varphi_0)$,
with analogous expressions for $\gamma$ and $F$,
\begin{align*}
    \partial_t \psi
    &\approx
    \sigma^2 \varphi_0 r_0 \gamma_0' \Delta \smooth{\gamma} \psi
    + \sigma^2 r_0 \gamma_0 \Delta \psi
    + \varphi_0 F'_0 \smooth{F} \psi .
\end{align*}
Letting $\hat f(u) = \int e^{2\pi i u x} f(x) dx$ denote the Fourier transform,
\begin{equation}
	\label{fourier transform psi}
    \partial_t \hat \psi(u)
    \approx
	\left\{
	- u^2\sigma^2 \varphi_0 r_0 \gamma_0' \hat{\rho}_\gamma(u) 
	%e^{-\epsilon^2 u^2 /2}
        - u^2 \sigma^2 r_0 \gamma_0
	+ \varphi_0 F'_0 \hat{\rho}_F(u) 
	%e^{-\epsilon^2 u^2 /2} 
	\right\} \hat \psi(u) .
\end{equation}
In the simplest case, in which $\gamma$ is constant, so $\gamma_0'=0$,
this reduces to
\begin{equation} %\begin{align*}
	\label{simplest FT}
	\partial_t \hat \psi(u)
    \approx
    \left(
%	- u^2\sigma^2 \varphi_0 r_0 \gamma_0' \hat{\rho}_\gamma(u) 
	%e^{-\epsilon^2 u^2 /2}
        - u^2 \sigma^2 r_0 \gamma_0
	+ \varphi_0 F'_0 \hat{\rho}_F(u) 
	%e^{-\epsilon^2 u^2 /2} 
    \right) \hat \psi(u) .
\end{equation} %\end{align*}
If we take $\rho_F=p_{\epsilon^2}$, then 
$\hat{\rho}_F(u)=\exp(-2\pi\epsilon^2 u^2)$ and (recalling 
that $F_0'<0$) the term in brackets is always negative, and we recover
the well-known fact that in this case the constant solution is stable.
If, on the other hand, $\hat{\rho}_F$ changes sign, there may be values of $u$ for which the corresponding quantity is positive. For example, if $d=1$ and
$\rho_F(x)={\mathbf 1}_{[-\epsilon, \epsilon]}(x)/2\epsilon$, then 
$\hat{\rho}_F(u)=\sin (2\pi\epsilon u)/(2\pi\epsilon u)$, which is 
negative for $u\in (1/(2\epsilon), 1/\epsilon)$ (and periodically 
repeating intervals). Setting $v=\epsilon u$, the bracketed term on 
the right hand side of~\eqref{simplest FT} becomes 
\[
	\varphi_0F_0'\frac{1}{2\pi v}\sin (2\pi v)
	-\frac{\sigma^2}{\epsilon^2}v^2r_0\gamma_0,
\]
and we see that if $\sigma^2/\epsilon^2$ is sufficiently small, there are
values of $v$ for which this is positive. In other words, in keeping
with our heuristic above, if dispersal is sufficiently short range relative to
the range over which individuals interact, there are unstable frequencies that
scale with the interaction distance $\epsilon$. In two dimensions, replacing 
the indicator of an interval by that of a ball of radius $\epsilon$, a
similar analysis applies, except that the sine function is replaced by a 
Bessel function.

Now suppose that $\gamma$ is not constant. Then, 
from~\eqref{fourier transform psi},
if we take 
$\rho_\gamma=\rho_F=p_\epsilon^2$,
\begin{equation*}
	\partial_t \hat \psi(u) = e^{-2\pi^2\epsilon^2 u^2 } 
	\left\{
		-\sigma^2\varphi_0 r_0\gamma_0'u^2
	-\sigma^2r_0\gamma_0u^2e^{2\pi^2\epsilon^2u^2}
	+\varphi_0 F_0'
	\right\}
	\hat{\psi}(u).
\end{equation*}
If we make the (reasonable)
assumption that $\gamma_0'<0$, then we see that even when the Fourier
transform of $\rho$ does not change sign, there may be parameter values
for which the constant solution is unstable.
As before, we set $v=\epsilon u$. The term in brackets becomes
\[
	\frac{\sigma^2}{\epsilon^2}v^2\left(-\varphi_0 r_0\gamma_0'
	-r_0\gamma_0e^{2\pi^2v^2}\right)+\varphi_0F_0',
\] 
and, provided $-\varphi_0\gamma_0'/\gamma_0>1$, for sufficiently small $v$ 
the term in round brackets is positive. We now see that if 
$\sigma^2/\epsilon^2$ is sufficiently {\em large},
the equilibrium state $\varphi\equiv\varphi_0$ 
is unstable. As before, the unstable frequencies will scale with $\epsilon$
and for given $F$, $r$ and $\gamma$, whether or not such unstable
frequencies exist will be determined by $\sigma^2/\epsilon^2$, but 
in this case of Gaussian kernels, 
it is interaction distance being sufficiently small relative
to dispersal that will lead to instability.

@@

We now look at the second order term in~(\ref{generator prelimit}).
An individual at location $x$ gives birth 
to a surviving offspring at $y$ at rate
$$
\gamma(x,\eta) r(y,\eta) q_{\theta}(x, dy) ,
$$
and since this increments $\langle f, \eta \rangle$ by $f(y) / N$,
the contribution to the quadratic variation from birth events,
which occur at rate $\theta$ per individual 
(so, rate $N\theta |\eta|$ overall), is
$$
\int_{\IR^d}
    N \theta \gamma(x,\eta)
    \int \frac{1}{N^2} f^2(y) r(y,\eta)
    q_\theta(x,dy) 
\eta(dx) .
$$
Similarly, the increment in $\langle f, \eta\rangle$ resulting from 
the death of an individual at $x$ is $-f(x)/N$, and so combining with the 
above, the second order term in the generator takes the form
\begin{align*}
& G''(\langle f,\eta\rangle)
\frac{1}{2} N \theta
\left\{
    \int_{\IR^d}
        \gamma(x,\eta)
        \int \frac{1}{N^2}f^2(y)r(y,\eta)q_\theta(x,dy) 
    \eta(dx)
    +
    \int_{\IR^d}
        \mu_\theta(x,\eta)\frac{1}{N^2}f^2(x) 
    \eta(dx)
\right\} \\
&\qquad
= \frac{1}{2} G''(\langle f, \eta \rangle)
    \frac{\theta}{N}
	\int_{\IR^d}\left\{
		\gamma(x, \eta) \int f^2(y) r(y, \eta) q_\theta(x, dy) + f^2(x) \mu_\theta(x, \eta) \right\}
    \eta(dx) .
\end{align*}
\comment{FROM ALISON integral against eta in last line corrected}
Since $\int f^2(y) r(y, \eta) q_\theta(x, dy) \to f^2(x) r(x, \eta)$
and $r\gamma  + \mu_\theta = 2 r \gamma  - F / \theta \to 2 r \gamma$
as $\theta \to \infty$,
this converges to
\begin{align*}
\frac{\alpha}{2} G''(\langle f, \eta \rangle)
    \big\langle
	2 r(x, \eta) \gamma(x, \eta),
        \eta(dx)
    \big\rangle .
\end{align*}

@@

Putting these together,
the level $u$ of an individual at $x$ evolves according to:
\begin{equation} \label{eqn:dot_u}
    \dot u
    =
    - \theta\frac{u}{N} \left(N - u\right)
    \gamma(x, \eta) \int_{\IR^d} r(y, \eta) q_\theta(x, dy) 
    +
    \theta\mu_\theta(x,\eta) u .
\end{equation}
We shall write 
\begin{align} \label{b_defn}
    b_\theta(x, \eta)
    :=
    \theta\left(
    \gamma(x,\eta) \int_{\IR^d} r(y, \eta) q_\theta(x, dy)
    -
    \mu_\theta(x,\eta)
    \right) ,
\end{align}
which captures the local net difference between reproduction and death,
and
\begin{align} \label{c_defn}
    c_\theta(x, \eta)
    :=
    \frac{\theta}{N} \gamma(x, \eta) \int_{\IR^d} r(y, \eta) q_\theta(x, dy) ,
\end{align}
which captures the local rate of production of successful offspring.
\comment{FROM ALISON This sort of true because it corresponds to lookdown events, but when I read this again I was confused. Is there a better way to say it?} 
Recall from equation~\eqref{eqn:mu_defn} that
$F(x,\eta) = \theta(r(x,\eta)\gamma(x,\eta) - \mu_\theta(x,\eta))$,
and so 
\begin{equation*} 
b_\theta(x, \eta)
=    \theta \gamma(x, \eta) \int_{\IR^d} \left( r(y, \eta) - r(x, \eta) \right) 
	q_\theta(x, dy)
    + F(x, \eta). 
\end{equation*}
Under Assumptions~\ref{def:model_setup}, as $\theta \to \infty$,
$c_\theta(x, \eta)$ will tend to $\alpha \gamma(x, \eta) r(x, \eta)$
and $b_\theta(x, \eta)$ will tend to
\begin{equation}
\label{eqn:b_limit}
    \gamma(x, \eta) \DG r(x, \eta) + F(x, \eta) .
\end{equation}

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:lookdown_mgale}
For given positive values of $N$ and $\theta$, define the generator $A^N$
by
\begin{equation}
	\label{eqn:lookdown_generator}
\begin{split}
& A^{N}f(\lp ) \\
&\quad =
    f(\lp)
    \,\sum_{(x,u)\in \lp}\,
    2 \gamma(x, \eta)
    \Bigg\{ \frac{1}{2N} \int_u^N g(x,u_1) du_1
            \frac{
                \theta \int_{\IR^d}
                (g(y,u) - g(x,u))
                r(y, \eta) q_{\theta}(x,dy)
            }{ g(x,u) }
        \\
    &\qquad\qquad\qquad\qquad\qquad\qquad\qquad {} +
        \frac{\theta}{N} \int_u^N
        \int_{\IR^d}\left(
            \frac{ g(y,u_1) + g(x,u_1) }{ 2 } - 1
        \right)
        r(y, \eta) q_{\theta}(x,dy)
        du_1
    \Bigg\}\\
    &\qquad\qquad {} +
    f(\lp) \sum_{(x,u)\in\lp}\,
    \left(
        c_\theta(x,\eta) u^2 - b_{\theta}(x,\eta)u
    \right)
    \frac{\partial_u g(x,u)}{g(x,u)}
    ,
\end{split}
\end{equation}
where $f(\lp) = \prod_{(x, u) \in \lp} g(x, u)$ is as defined in \eqref{eqn:test_functions},
and $\eta(\cdot) = \lp(\cdot \times [0, N]) / N$
as before. 
Given $\lp_0 \in {\mathcal C}_N$, 
we say that a ${\mathcal D}_{[0,\infty)}({\mathcal C}_N)$-valued process 
	\comment{FROM ALISON: we have not been consistent in the layout of 
	the notation for spaces of cadlag paths - it is already different in 
	the definition below!}
$(\xi_t)_{t \ge 0}$
   is a solution to the $(A^N, \lp_0)$ martingale problem
    if 
$f(\lp_t) - f(\lp_0)-\int_0^t A^N f(\lp_s) ds$ is a martingale (with 
respect to the natural filtration) for all 
test functions $f$ 
    as defined above.
\end{definition}

@@

To make this precise,
define the spatial projection maps
$\kappa^N: \mathcal{M}(\IR^d \times [0, N]) \to \mathcal{M}(\IR^d)$
by $\kappa^N(\xi^N)(\cdot) = \xi^N(\cdot \times [0, N]) / N$,
and $\kappa: \mathcal{M}(\IR^d \times [0, \infty)) \to \mathcal{M}(\IR^d)$
by $\kappa(\xi)(\cdot) = \lim_{u_0 \to \infty} \xi(\cdot \times [0, u_0]) / u_0$.
We will also need an inverse notion:
for a measure $\lp^N$ on $\IR^d \times [0, N]$ and a $\sigma$-field $\mathcal{F}$,
we say that \emph{$\lp^N$ is conditionally uniform given $\mathcal{F}$}
if for all compactly supported $f$,
\begin{align} \label{eqn:conditionally_uniform}
    \IE[ e^{-\langle f, \lp \rangle} \;|\; \mathcal{F} ]
    &=
    e^{-\langle H_f^N, \kappa^N(\lp) \rangle} ,
\end{align}
where
$$ H_f^N(x) = - N \log \frac{1}{N} \int_0^N e^{-f(x, u)} du . $$
In other words, the $[0, N]$ components of $\lp$
are independent, uniformly distributed on $[0, N]$, and independent of 
$\kappa^N(\lp)$.
\comment{ALISON: added N, also made sigma $\sigma$ above}
Similarly, for a measure $\lp$ on $\IR^d \times [0, \infty)$
we say that \emph{$\lp$ is a conditionally Poisson random measure given $\mathcal{F}$}
if for all compactly supported $f$,
\begin{align} \label{eqn:conditionally_poisson}
    \IE[ e^{-\langle f, \lp \rangle} \;|\; \mathcal{F} ]
    &=
    e^{-\langle \int_0^\infty (1 - e^{-f(x, u)}) du, \kappa(\lp)(dx) \rangle} .
\end{align}
In other words, $\lp$ is conditionally Poisson with Cox measure $\kappa(\lp) \times \lambda$,
where $\lambda$ is Lesbegue measure.

\begin{proposition}
    \label{thm:mmt_application}
    If $\tilde \eta^N$ is a solution
    of the martingale problem of Definition~\ref{defn:mgale_construction}
    with initial distribution $\eta^N_0$
    then there exists a solution $\lp^N$
    of the $(A^N, \lp^N_0)$-martingale problem of Definition~\ref{defn:lookdown_mgale}
    such that $\eta^N = \kappa^N \circ \lp^N$
    has the same distribution on $D_{\measures}[0, \infty)$
    as $\tilde \eta^N$.
    Furthermore, for each $t$, $\lp^N_t$ is conditionally uniform 
	\comment{ALISON deleted excess given}
	given $\mathcal{F}^{\eta^N}_t$
    in the sense of \eqref{eqn:conditionally_uniform}.
